Variance reduction for MCMC methods via martingale representations

### 2020-04-12 aap/Revision/vr-mcmc-revision-new-new.tex
update

Monte Carlo and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian Statistics. Suppose that we aim at computing $\pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$. Typically it is not possible to compute $\pi(f)$ analytically, leading to the Monte Carlo methods as a popular solution. Given an independent identically distributed observations $X_1,\ldots,X_n$ form $\pi$, we might estimate $\pi(f)$ by $\hat{\pi}(f):=1/n\sum_{k=1}^{n}f(X_k)$. The variance of such estimate equals $\sigma^2(f)/n$  with $\sigma^2(f)$ being the variance of the integrand. The first way to obtain tighter estimate $\hat{\pi}(f)$ is simply to increase the sample size $n$. Unfortunately, this solution might be prohibitively costly, especially when the dimension $d$ is large enough or sampling from $\pi$ is complicated. An alternative approach is to decrease $\sigma^2(f)$ by constructing a new Monte-Carlo experiment with the same expectation as the original one, but with a lower variance. Methods to achieve this are known as variance
reduction techniques. Introduction to many of them can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
\par

One of the popular approaches to variance reduction is the so called control variates method. We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v. $f(X) - \zeta$ is small. The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
Markov chain \((X_p),\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows. One important class of the variance reduction methods in this case  is based on the Poisson equation for the corresponding  Markov chain. It was observed in~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).  The best choice for the function \(G\) corresponds to a solution of the Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).  Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.   Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.
\par
In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
\par
The paper is organized as follows.  In Section~\ref{sec:setup} we set up the problem and introduce some notations. Section~\ref{seq:mart_repr} contains a novel martingale representation and shows how this representation can be used for variance reduction. In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA).  Finally, numerical examples are presented in Section~\ref{sec:num}.

@@

\subsection{Comparison with vanilla ULA}
In this subsection we aim at comparing cost of variance reduction, achieved by the proposed algorithm, to the cost of vanilla ULA algorithm. We consider samples, generated by ULA with $\pi$ being either the standard normal distribution in dimension $d$ or the mixture of two $d-$dimensional standard Gaussians of the form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-\frac{\|x-\mu\|^2}{2}} + \rme^{-\frac{\|x+\mu\|^2}{2}}\right)
 \]
 where $d = 2$ and $\mu = (0.5,0.5)$. For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$. We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order polynomial approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively. Then we estimate the cost-variance ratio
\begin{eqnarray*}
\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}
\end{eqnarray*}
by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$. For different $n_{\mathrm{trunc}}$ and $K$ we estimate $\mathrm{cost}(\widehat \pi_{K,n}^N)$ as 
 \begin{equation}
 \label{eq:cvR}
 %check this
 %add n_{truncation} here
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times d^{K}
 \end{equation}
 since the representations of $a_{r,\mathbf{k}} (x)$ might be pre-computed using formulas \eqref{eq:a_hat}. Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}. Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.

>>>>

Monte Carlo and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian statistics. Suppose that we aim at computing $\pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$. Typically it is not possible to compute $\pi(f)$ analytically, leading to the Monte Carlo methods as a popular solution. Given an independent identically distributed observations $X_1,\ldots,X_n$ form $\pi$, we might estimate $\pi(f)$ by $\hat{\pi}(f):=1/n\sum_{k=1}^{n}f(X_k)$. The variance of such estimate equals $\sigma^2(f)/n$  with $\sigma^2(f)$ being the variance of the integrand. The first way to obtain tighter estimate $\hat{\pi}(f)$ is simply to increase the sample size $n$. Unfortunately, this solution might be prohibitively costly, especially when the dimension $d$ is large enough or sampling from $\pi$ is complicated. An alternative approach is to decrease $\sigma^2(f)$ by constructing a new Monte-Carlo experiment with the same expectation as the original one, but with a lower variance. Methods to achieve this are known as variance
reduction techniques. Introduction to many of them can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
One of the popular approaches to variance reduction is the so called control variates method. We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v. $f(X) + \zeta$ is small. The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\PE[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.
Unfortunately it is not always possible to generate independent observations distributed according to $\pi$. To overcome this issue one might consider MCMC algorithms, where the exact samples from $\pi$ are replaced by \((X_p),\) \(p=0,1,2,\ldots,\) forming a Markov chain with a marginal distribution of $X_n$ converging to $\pi$ in a suitable metrics as $n$ grows. It is still possible to apply the control variates method in the similar manner to a simple Monte-Carlo case, yet the choice of the optimal control variate becomes much more difficult. Due to significant correlations between the elements of the Markov chain it might be not enough to minimize the marginal variances of \((X_p)_{p \geq 0}\) as it was in independent case. Instead one may choose the control variate by minimizing the corresponding asymptotic variance of the chain as it is suggested in~\cite{belomestny2019esvm}. At the same time it is possible to express the optimal control variate in terms of the solution of the Poisson equation for the corresponding Markov chain \((X_p)\). As it was observed in~\cite{henderson1997variance}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has a zero mean with respect to $\pi$ for arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\). Hence, $U_{G}(x)$ is a valid control functional for a suitable choice of $G$, with the best $G$ given by a solution of the Poisson equation \(\PE [G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\). For such $G$ we obtain $f(x) + U_G(x) = f(x) - f(x) + \pi(f) = \pi(f)$ leading to an ideal estimator with zero variance. Despite the fact that the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.
In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
The paper is organized as follows.  In Section~\ref{sec:setup} we set up the problem and introduce some notations. Section~\ref{seq:mart_repr} contains a novel martingale representation and shows how this representation can be used for variance reduction. In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA).  Finally, numerical examples are presented in Section~\ref{sec:num}.

@@

\subsection{Comparison with vanilla ULA}
In this subsection we aim at comparing cost of variance reduction, achieved by the proposed algorithm, to the cost of vanilla ULA algorithm. We consider samples, generated by ULA with $\pi$ being either the standard normal distribution in dimension $d$ or the mixture of two $d-$dimensional standard Gaussians of the form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-\frac{\|x-\mu\|^2}{2}} + \rme^{-\frac{\|x+\mu\|^2}{2}}\right)
 \]
 where $d = 2$ and $\mu = (0.5,0.5)$. For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$. We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order polynomial approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively. Then we estimate the cost-variance ratio
\begin{eqnarray}
\label{eq:cvR}
\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
\end{eqnarray}
by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$. Since for fixed $K$ the cost of computing $\pi_{n}^N(f)$ is proportional to computing function $f$, we set for $K = 1$
 \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 2
 \end{equation*}
 since there are only $2$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$. Note that in this case each $a_{r,\mathbf{k}}(x)$  is a linear function, which can be computed at the same cost as $f$. Similarly, for $K = 2$ we set
  \begin{equation*}
 \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 8
 \end{equation*}
since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$. Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}. Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.
